Refining from the VHDL-AMS Hardware Description Language to the BON High-level Specification Language

### 2009-09-17 papers/BON_VHDL/paper.tex
first draft

\lstset{
  language=Java,
  basicstyle=\footnotesize\sffamily,
  identifierstyle=\itshape,
  stringstyle=\footnotesize\ttfamily,
  commentstyle=\textup,
  columns=flexible,
  captionpos=b,
  morekeywords={},
  literate=
  {-}{{$\textendash$}}1
  {<}{{$\textless$}}1
  {<=}{{$\textless=$}}1
  {>}{{$\textgreater$}}2
  {>=}{{$\textgreater=$}}2
  {\\}{{$\backslash\backslash$}}1
  {'}{{$\textprimstress$}}1
}

@@

\lstdefinelanguage{Bon} {
  morekeywords={require,ensure,feature,old,and,o,Q'slew,end,Result, Void},
  morekeywords={class_chart,class,delta,invariant,inherit,deferred,effective,NONE,make},
  morekeywords={explanation,query,command,constraint,exists},
  morecomment=[l]{--},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
 literate=,
  showstringspaces=false
}

\lstdefinelanguage{VHDL} {
  morekeywords={not,xnor,and,nand, nor,or,xor,integer,byte,real},
  morekeywords={character,enum,type,boolean,bit,time,null,std_ulogic},
  morekeywords={sll, srl, sla, sra, rol, ror},
  morekeywords={rem,abs,mod},
  morekeywords={terminal,T'reference, T'contribution,T'tolerance},
  morekeywords={reference,contribution},
  morekeywords={nature,subnature,electrical,magnetic,translational,translational_v},
  morekeywords={rotational,rotational_v,fluidic,thermal,radiant},
  morekeywords={voltage,angular_velocity,pressure,velocity},
  morekeywords={current,force,torque,volumetric,flow_rate},
  morekeywords={T'left,T'right,T'high,T'ascending,T'image(x),T'value(s)},
  morekeywords={left,right,high,ascending,image,value},
  morekeywords={T'pos(s),T'val(x),T'succ(x),T'pred(x)},
  morekeywords={pos,val,succ,pred},
  morekeywords={A'left(n),A'right(n),A'low(n),A'high(n)},
  morekeywords={left,right,low,high},
  morekeywords={A'range(n),A'reverse_range(n),A'length(n),A'ascending(n)},
  morekeywords={range,reverse_range,length,ascending},
  morekeywords={T'leftof(x),T'rightof(x),T'base},
  morekeywords={leftof,rightof,base},
  morekeywords={Q'tolerance, Q'above,Q'delayed,Q'dot,Q'integ},
  morekeywords={delayed,dot,integ},
  morekeywords={Q'slew,Q'ztf,Q'ltf,Q'zoh},
  morekeywords={N'across, N'through},
  morekeywords={across,through,zoh,ltf,slew,ztf},
  morekeywords={above,tolerance,group,quantity},
  morekeywords={entity,end,use,port,generic,is,library,all},
  morekeywords={architecture,of,through,begin,constant},
  morecomment=[l]{--},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
 literate=,
  showstringspaces=false
}

@@

%======================================================================
\section{Introduction}
\label{sec:introduction}
The \ST sensor platform, represents heterogeneous 
collection of digital, analog and mixed signal hardware circuits. 
The complexity of our platform requires the need to strive for formal 
descriptions each component separately and  behavior of their 
complex interdependencies, in order to produce executable specification.
To make sure that the sensor platform works properly we need 
to translate the VHDL-AMS code into the specification, and to refine and 
verify this specification.
In other words, our aim is to establish formal link between behavior of a given 
sensor platform, and derived specification of components  in order to implement 
fully validated and verified \ST simulator.
For this purpose, we need notation and methodology 
which will be able to satisfy the requirements given below.\\ 
First, both of them, should be in sync with  VHDL-AMS language and reference 
manual which tries to define VHDL-AMS as well as possible in a descriptive way,
explaining the semantics in English.\\
Second, specification derived from the notation should  be easy to attach to 
the rest of the system architecture\xspace(driver, microprocessor emulator, 
test-benches e.t.c.).\\
Third, methodology must aid efforts on creation of target system 
correctness criteria, such that are independent of implementation.
It means that verification of simulator model should take place on specification 
using reasoning techniques provided for algebraic formalism before coding starts.\\
Fourth,a precise semantics should  provides the basis for
automated forward-backward engineering.\\
Managing these requirements, and insuring that the platform definition 
(constraints) are preserved and verified through all levels of 
abstraction is where one becomes aware of gaps in already 
used design methodologies.
In our opinion, the BON-JML approach is in line with requirements given above.
The authors will illustrate in this article how BON can represent a high
level analogue mapping notation and will further illustrate a refinement 
technique from VHDL-AMS to BON and JML.\\
This article provides three main contributions.
\begin{enumerate}
\item
 It provides a solution how to formally emulate
 piece of hardware using BON-JML approach; up to date BON and JML
 have been used mostly in the pure object oriented software construction.   
\item
It relates hardware description code and refinement technique.
It means that if a satisfactory VHDL-AMS code exists, then it can be simulated  from the 
BON (informal/formal) specification, and JML implementation by an 
appropriate sequence of refinements.
\item
It introduces \inv as a general concept of modeling hardware 
components behavior and their interactions.
To be more precisely, the behavioral specification of the \STS may 
be represented as a \inv on the signals at the circuit's ports and 
may be obtained by taking the conjunction of the \emph{invariants} 
that describe behaviors of the individual components separately.\\ 
\end{enumerate}
The rest of this article is organized as follows.
In \autoref{sec:vhdlams} we have  presented an overview of VHDL-AMS language.
\autoref{sec:choiceobon} defines  BON principles and explains the
reasons why we have decided to use BON method.
In \autoref{sec:mappingVB} we will show mapping between VHDL-AMS and BON. 
The central part of the our article is \autoref{sec:refVB}, in which
we will give practical aspect of refinement , applied on 
first order passive RC Low-Pass filter, in order to produce formal 
specification and software simulation.
In \autoref{sec:conclusion} the conclusions are given,
and directions of future studies are outlined. Finally, a comprehensive list 
of the mapping between BON and VHDL-AMS can be found in the appendix. \\ 
%=====================================================================

@@

\section{Refinement from VHDL-AMS to BON }
\label{sec:refVB}
The main theme of this article is the translation from VHDL-AMS to BON using 
refinement approach.
We will refine the VHDL-AMS code up to the formal BON, and from this 
specification we will generate an executable specification in the JML language.
For this purpose, we will demonstrate our approach in BON-JML fashion trought 
low-pass filter example.
\input{lowpassRC}
According to equation given above (see \autoref{fig:lowpassfig}) 
we can produce VHDL-AMS \emph{simultaneous statement} using 
\texttt{$s \rightarrow d/dt \rightarrow 'dot$} relation.
The transformation looks as follows: \\
\begin{equation}
v_{in} = v_{out} + \tau_psv_{out} 
   \begin{array}[b]{c}                                                                                                                                    
      {\scriptscriptstyle s \rightarrow d/dt}  \\=                                                                                                                  
   \end{array} 
v_{out} + \tau_p\cfrac{dv_{out}}{dt}
\begin{array}[b]{c}                                                                                                                                    
      {\scriptscriptstyle d/dt \rightarrow } {\scriptscriptstyle 'dot} \\=                                                                                                                  
   \end{array}
  v_{in} = v_{out} + \tau_pv_{out}{'dot} 
\end{equation} 
\input{lowpass}
We have not yet said what specification and refinement are.
Let us introduce these terms first.\\ 
A specification of a program component is a contract that gives some constraints 
on how the component is to behave but leaves freedom for the implementator to 
decide how the actual behavior of the component is to be realized.
Specification involves \emph{choice} between a number of alternatives; this 
 relationship is one-to-many. As we have shown in \autoref{tab:rescap}, corresponding 
informal specifications represents high level of abstraction derived from VHDL-AMS code.

A refinement represents the concept that can be understud as series of 
correctness-preserving transformations from general perspective to 
a precise level of detail. Accurately, if we compare two specifications,
\texttt{S} and \texttt{S'}, then we can conclude that the \texttt{S} 
is more abstract than \texttt{S'} if any condition that we can establish 
with the first specification can also be established by second. 
It means that "\texttt{S is \emph{refined} by S'}", 
and we denote this  as \texttt{S $\rightsquigarrow$ S'}.
Let say that the first informal BON specification is denoted by 
\texttt{S} (see \autoref{tab:rescap}), and second, which is given in Table2  is denoted by \texttt{S'}.
In this case, the refinement relation represents an \emph{isomorphism} 
and maps the following:
%=====================================================================
\section{Conclusion}
\label{sec:conclusion}
The BON notation has originally been used for description of 
software models. However, after certain elaboration it can be used 
for description of hardware components. Practical utilization of the approach
in the project on mapping between VHDL-AMS and BON using refinement technique,
has demonstrated through low-pass filter example and shown that simulation
of component requires relatively little and inexpensive  efforts. Moreover, our article shown
that the development of the specification produced by the BON and translated 
into JML can be applied as refinement procedure: from VHDL-AMS to informal BON, 
next, from informal to more formal BON and finally, to JML specification. 
The BON support general software development and should be adapted for concrete
applications and language as VHDL-AMS. For this purpose we have developed various 
VHDL-AMS related types, operators and relations and we have defined appropriate 
specifications for all of them. However, the behavior of
\texttt{break} statement and  concurrency, for example, have not been translated 
yet in the BON notation. We have started a new phase of project named \emph{eBon} in which
the issues mentioned above, and creation of appropriate components and relations 
are main subject.

@@

%======================================================================
\appendix
\newpage
\section {APPENDIX}
\input{static}
\input{arith}
\input{shift}
\input{relation}
\input{logic}
\input{types}
\input{assignment}
\input{iscaltypeattr}
\input{tolerance}
\input{effort}
\input{flow}
\input{nature}
\input{iscalnature}
\input{iarrayattr}
\input{scalnat}
\input{iterminal}
\input{terminal}
\input{iquantity}
\input{quantity}
\input{quantityport}
\input{terminalport}
\input{signal}
\input{signalattr}
\input{wait}
\input{bitbon}
%======================================================================

>>>>

\lstset{
  language=Java,
  basicstyle=\footnotesize\sffamily,
  identifierstyle=\itshape,
  stringstyle=\footnotesize\ttfamily,
  commentstyle=\textup,
  columns=flexible,
  captionpos=b,
  morekeywords={},
  literate=
  {-}{{$\textendash$}}1
  {<}{{$\textless$}}1
  {<=}{{$\textless=$}}1
  {>}{{$\textgreater$}}2
  {>=}{{$\textgreater=$}}2
  {'}{{$\textprimstress$}}1
}

@@

\lstdefinelanguage{Bon} {
  morekeywords={require,ensure,feature,old,and,o,Q'slew,end,Result, Void},
  morekeywords={class_chart,class,delta,invariant,inherit,deferred,effective,NONE,make},
  morekeywords={explanation,query,command,constraint,exists,for_all,such_that,it_holds},
  morecomment=[l]{--},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
 literate=,
  showstringspaces=false
}

\lstdefinelanguage{VHDL} {
  morekeywords={not,xnor,and,nand, nor,or,xor,integer,byte,real},
  morekeywords={character,enum,type,boolean,bit,time,null,std_ulogic},
  morekeywords={sll, srl, sla, sra, rol, ror},
  morekeywords={rem,abs,mod,to},
  morekeywords={terminal,T'reference, T'contribution,T'tolerance},
  morekeywords={reference,contribution},
  morekeywords={nature,subnature,electrical,magnetic,translational,translational_v},
  morekeywords={rotational,rotational_v,fluidic,thermal,radiant},
  morekeywords={voltage,angular_velocity,pressure,velocity},
  morekeywords={current,force,torque,volumetric,flow_rate},
  morekeywords={T'left,T'right,T'high,T'ascending,T'image(x),T'value(s)},
  morekeywords={left,right,high,ascending,image,value},
  morekeywords={T'pos(s),T'val(x),T'succ(x),T'pred(x)},
  morekeywords={pos,val,succ,pred},
  morekeywords={A'left(n),A'right(n),A'low(n),A'high(n)},
  morekeywords={left,right,low,high},
  morekeywords={A'range(n),A'reverse_range(n),A'length(n),A'ascending(n)},
  morekeywords={range,reverse_range,length,ascending},
  morekeywords={T'leftof(x),T'rightof(x),T'base},
  morekeywords={leftof,rightof,base},
  morekeywords={Q'tolerance, Q'above,Q'delayed,Q'dot,Q'integ},
  morekeywords={delayed,dot,integ},
  morekeywords={Q'slew,Q'ztf,Q'ltf,Q'zoh},
  morekeywords={N'across, N'through},
  morekeywords={across,through,zoh,ltf,slew,ztf},
  morekeywords={above,tolerance,group,quantity},
  morekeywords={entity,end,use,port,generic,is,library,all},
  morekeywords={architecture,of,through,begin,constant},
  morecomment=[l]{--},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
 literate=,
  showstringspaces=false
}

@@

%======================================================================
\section{Introduction}
\label{sec:introduction}
The \ST sensor platform, represents heterogeneous 
collection of digital, analog and mixed signal hardware circuits. 
The complexity of our platform requires the need to strive for formal 
descriptions each component separately and  behavior of their 
complex interdependencies, in order to produce executable specification.
To make sure that the sensor platform works properly we need 
to translate the VHDL-AMS code into the specification, and to refine and 
verify this specification.
In other words, our aim is to establish formal link between behavior of a given 
sensor platform, and derived specification of components  in order to implement 
fully validated and verified \ST simulator.
For this purpose, we need notation and methodology 
which will be able to satisfy the requirements given below.\\ 
First, both of them, should be in sync with  VHDL-AMS language and reference 
manual which tries to define VHDL-AMS as well as possible in a descriptive way,
explaining the semantics in English.\\
Second, specification derived from the notation should  be easy to attach to 
the rest of the system architecture\xspace(driver, microprocessor emulator, 
test-benches e.t.c.).\\
Third, methodology must aid efforts on creation of target system 
correctness criteria, such that are independent of implementation.
It means that verification of simulator model should take place on specification 
using reasoning techniques provided for algebraic formalism before coding starts.\\
Fourth, a precise semantics should  provides the basis for
automated forward-backward engineering.\\
Managing these requirements, and insuring that the platform definition 
(constraints) are preserved and verified through all levels of 
abstraction is where one becomes aware of gaps in already 
used design methodologies.
In our opinion, the BON-JML approach is in line with requirements given above.
The authors will illustrate in this article how BON can represent a high
level analogue mapping notation and will further illustrate a refinement 
technique from VHDL-AMS to BON and JML.\\
This article provides three main contributions.
\begin{enumerate}
\item
 It provides a solution how to formally emulate
 piece of hardware using BON-JML approach; up to date BON and JML
 have been used mostly in the pure object oriented software construction.   
\item
It relates hardware description code and refinement technique.
It means that if a satisfactory VHDL-AMS code exists, then it can be simulated  from the 
BON (informal/formal) specification, and JML implementation by an 
appropriate sequence of refinements.
\item
It introduces \inv as a general concept of modeling hardware 
components behavior and their interactions.
To be more precisely, the behavioral specification of the \STS may 
be represented as a \inv on the signals at the circuit's ports and 
may be obtained by taking the conjunction of the \emph{invariants} 
that describe behaviors of the individual components separately.\\ 
\end{enumerate}
The rest of this article is organized as follows.
In \autoref{sec:vhdlams} we have  presented an overview of VHDL-AMS language.
\autoref{sec:choiceobon} defines  BON principles and explains the
reasons why we have decided to use BON method.
In \autoref{sec:mappingVB} we will show mapping between VHDL-AMS and BON. 
The central part of the our article is \autoref{sec:refVB}, in which
we will give practical aspect of refinement, applied on 
first order passive RC low-pass filter, in order to produce formal 
specification and software simulation. In this section we 
explore this approach in detail.
In \autoref{sec:conclusion} the conclusions are given,
and directions of future studies are outlined. Finally, a comprehensive list 
of the mapping between BON and VHDL-AMS can be found in the appendix. \\ 
%=====================================================================

@@

\section{Refinement from VHDL-AMS to BON }
\label{sec:refVB}
The main theme of this article is the translation from VHDL-AMS to BON using 
refinement approach.
We will refine the VHDL-AMS code up to the formal BON, and from this 
specification we will generate an executable specification in the JML language.
For this purpose, we will demonstrate our approach in BON-JML fashion trought  
low-pass filter example.\\
\subsection{Low-Pass filter transfer function }
\label{subsec:lptransfunc} 
\input{lowpassRC}
The most simple low-pass filter is passive RC network 
shown in \autoref{fig:lowpassfig}.
According to equation given above (see \autoref{fig:transfun}) 
we can produce VHDL-AMS \emph{simultaneous statement} using 
\texttt{$s = d/dt = 'dot$} relation.
The transformation looks as follows: \\
\begin{equation}
v_{in} = v_{out} + \tau_psv_{out} 
   \begin{array}[b]{c}                                                                                                                                    
      {\scriptscriptstyle s = d/dt}  \\=                                                                                                                  
   \end{array} 
v_{out} + \tau_p\cfrac{dv_{out}}{dt}
\begin{array}[b]{c}                                                                                                                                    
      {\scriptscriptstyle d/dt = } {\scriptscriptstyle 'dot} \\=                                                                                                                  
   \end{array}
  v_{in} = v_{out} + \tau_pv_{out}{'dot} 
\end{equation} 
 The last equation can be described as follows. 
The input voltage\xspace($v_{in}$)\xspace equals summation 
of output voltage\xspace($v_{out}$)\xspace and multiplication 
of the time constant\xspace($\tau_p$)\xspace of the pole and the
first derivative of the capacitor's voltage\xspace($v_{out}{'dot}$)\xspace 
over time.\\
Now we can discretize previous equation. Let us represent $v_{in}$ by a series 
of samples $v_{in(1..n)}$. At the same points in time, we can represent $v_{out}$ by 
series of sample $v_{out(1..n)}$. We can assume that samples are taken at points in 
time incrementing by $\triangle$t. Now we have equivalent equations:
\begin{equation}
v_{in(i)}- v_{out(i)}=\tau_p(\frac{v_{out(i)}- v_{out(i-1)}} {\triangle t});                                                                                                                  
\end{equation}

\begin{equation}
v_{out(i)}=v_{in(i)}(\frac{\triangle t}{\tau_p + \triangle t}) +
 v_{out(i-1)}(\frac{\tau_p}{\tau_p + \triangle t});                                                                                                                  
\end{equation}

\begin{equation}
v_{out(i)} 
 \begin{array}[b]{c}                                                                                                                                    
      {\scriptscriptstyle \alpha= \triangle t / (\tau_p + \triangle t)}  \\=                                                                                                           
 \end{array}
 \alpha v_{in(i)} + (1- \alpha) v_{out(i-1)};                                                                                                                  
\end{equation}

The last equation gives us a way to calculate the output samples in terms of the 
input samples and the preceding output.
But how to represent these unknowns in BON ?
We introduce the  \lstinlinen|BRANCH_QUANTITY| objects like wrappers around unknowns,
with features as fallows:
\begin{itemize}
\item
\emph{across} which represent current:$i_{out}$ 
\item 
\emph{through} which represent voltage: $v_{in}$, $v_{out}$. 
\item 
\emph{terminal sequences} which describes input/output terminal(s). 
\end{itemize}
Now we are in presence of BON notation and we may start
with the  refinement of the VHDL-AMS code.\\
\subsection{Refine example }
\label{subsec:refine}     
We have not yet said what specification and refinement are.
Let us introduce these terms first.\\
A \emph{specification} of a program component is a 
contract that gives some constraints 
on how the component is to behave but leaves freedom for the implementator to 
decide how the actual behavior of the component is to be realized. 
As we have shown in \autoref{tab:rescapfirst}, corresponding 
informal specifications represents high level of abstraction derived 
from VHDL-AMS code.\\
A \emph{refinement} represents the concept that can be understood as series of 
correctness-preserving transformations from general perspective to 
a precise level of detail. Accurately, if we compare two specifications,
\texttt{S} and \texttt{S'}, then we can conclude that the \texttt{S} 
is more abstract than \texttt{S'} if any condition that we can establish 
with the first specification can also be established by second. 
It means that "\texttt{S is \emph{refined} by S'}", 
and we denote this  as \texttt{S $\sqsubseteq$ S'}.
The \emph{refinement} which was derived in three steps, 
from informal to more formal BON and finally to JML, 
is represented using relation \texttt{$\sqsubseteq$}. This relation represent 
\emph{isomorphism} and has inverse relation \texttt{$\sqsupseteq$}. 
It means that once we have more formal specification, we can refine 
and seamlessly  translate specification in previous abstract level,  
keeping specification validity and documentation consistency. 
In other  words, we use two important BON principles to achieve 
as much as precise model consistency: the seamlessness and  
reversibility(see \autoref{sec:choiceobon}). 
Let say that the first informal BON specification is denoted by 
\texttt{S} (see \autoref{tab:rescapfirst}), and second and third,
which are given in \autoref{tab:rescapsecond}  is denoted by \texttt{S',S''}.
The forward process refinement can be represented as:\xspace
\texttt{S $\sqsubseteq$ S' $\sqsubseteq$ S''}, while the backward process
can be represented as \xspace \texttt{S''$\sqsubseteq$ S' $\sqsubseteq$ S}.  
Using all BON principles, we have developed an specification of RC low-pass
filter, and we have produced fully validated and verified low-pass filter.
\input{lowpass1}
\input{lowpass2}

%=====================================================================
\section{Conclusion}
\label{sec:conclusion}
The BON notation has originally been used for description of 
software models. However, after certain elaboration it can be used 
for description of hardware components. Practical utilization of the approach
in the project on mapping between VHDL-AMS and BON using refinement technique,
has demonstrated through low-pass filter example and shown that simulation
of component requires relatively little and inexpensive  efforts. Moreover, our article shown
that the development of the specification produced by the BON and translated 
into JML can be applied as refinement procedure: from VHDL-AMS to informal BON, 
next, from informal to more formal BON and finally, to JML specification. 
The BON support general software development and should be adapted for concrete
applications and language as VHDL-AMS. For this purpose we have developed various 
VHDL-AMS related types, operators and relations and we have defined appropriate 
specifications for all of them. However, the behavior of
\texttt{break} statement and  concurrency, for example, have not been translated 
yet in the BON notation. We have started a new phase of project named \emph{eBon} in which
the issues mentioned above, and creation of appropriate components and relations 
are main subject.

@@

%======================================================================
\appendix
\newpage
\section {APPENDIX}
\input{static}
\input{arith}
\input{shift}
\input{relation}
\input{logic}
\input{types}
\input{assignment}
\input{iscaltypeattr}
\input{tolerance}
\input{effort}
\input{flow}
\input{nature}
\input{iscalnature}
\input{iarrayattr}
\input{scalnat}
\input{iterminal}
\input{terminal}
\input{iquantity}
\input{quantity}
%\input{quantityport}
%\input{terminalport}
%\input{signal}
%\input{signalattr}
%\input{wait}
%\input{bitbon}
%======================================================================
